### 2016-02-29 degeneracy/apps/degen_magnitude.Rmd
apps and restructure

---
title: "Degeneracy and parameter magnitude"
author: "Andee Kaplan"
date: "March 1, 2015"
output:
  html_document:
    theme: readable
runtime: shiny
---

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(ggvis)
library(ggplot2)
library(dplyr)
library(tidyr)
library(shiny)
```

@@

plot_data <- function(res) {

  plot.data <- data.frame()
  for(i in 1:nrow(res)) {
    tmp <- res$samp[[i]]
    H <- res[i,]$H
    V <- res[i,]$V

    tmp %>% 
      data.frame() %>% 
      rowwise() %>% 
      mutate_(ss_ratio = paste0("(", paste(paste0(names(.)[(H + V + 1):(H + V + H*V)], "^2"), collapse = " + "), ")/sum(", paste(paste0(names(.)[1:(H+V)],"^2"), collapse = " + "), ")")) %>%    
      ungroup() -> ratio

    inner_join(ratio, res$outside[[i]] %>% ungroup()) %>%
      select(ss_ratio, near_hull) %>%
      mutate(H = H, V = V, n_param = H + V + H*V) %>%
      rbind(plot.data) -> plot.data
  }
  return(plot.data)
}

plot.dat <- list()
for(i in seq(0, 1, by = 0.1)) {
  load(paste0("results_", i, ".RData")) 
  plot_dat <- lapply(res, plot_data)
  names(plot_dat) <- gsub("[.]", "_", r_center)
  plot.dat[[i*10 + 1]] <- plot_dat
}

plot.dat_split <- list()
for(i in seq(0, 1, by = 0.1)) { #full data not compiled yet
  load(paste0("results_split_", i, ".RData")) 
  res <- lapply(res, function(x) x %>% mutate(r = r2)) #dumb naming convention
  plot_dat_split <- lapply(res, plot_data)
  names(plot_dat_split) <- gsub("[.]", "_", r_center)
  plot.dat_split[[i*10 + 1]] <- plot_dat_split
}

@@

# ggvis.subset %>%
#   ggvis(~n_param_f, ~ss_ratio) %>%
#   layer_points(fill = ~as.character(near_hull), opacity := 0.4, size := 200) %>%
#   add_axis("x", title = "# Parameters") %>%
#   add_axis("y", title = "Ratio of SS of cross terms to main effects") %>%
#   add_legend("fill", title = "Within 0.05 of Hull")

>>>>

---
title: "Degeneracy and parameter magnitude"
output:
  html_document:
    theme: readable
runtime: shiny
---

```{r libraries, message=FALSE, warning=FALSE, echo=FALSE}
library(ggvis)
library(ggplot2)
library(dplyr)
library(tidyr)
library(shiny)
source("functions.R")
```

@@

load("plot_dat.RData")
```

@@

### 2016-02-29 presentations/coda2016/coda2016_poster.tex
apps and restructure

\begin{table}
\centering
\begin{tabular}{| p{.48\linewidth} | p{.5\linewidth} |}
\hline
Characteristic & Detection \\
\hline
\hline
{\bfseries Near-degeneracy.} Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model \cite{handcock2003assessing}. & If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$. \\
\hline
{\bfseries Instability.} Small changes in natural parameters result in large changes of the pmf, excessive sensitivity \cite{schweinberger2011instability}. & If for any $C > 0$ there exists $N_C > 0$ such that 
$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$. Where $Q(\cdot)$ is the neg-potential function of the model. \\
\hline
{\bfseries Uninterpretability.} Due to the existence of dependence, marginal mean-structure no longer maintained \cite{kaiser2007statistical}. & If the magnitude of the difference between model expectations and expectations under independence, $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
\hline
\end{tabular}
\label{tab:degen}
\caption{Table of ``improper model'' characteristics.}
\end{table}
\end{textblock}

%%-----------------------------------------------
%%
%%Bottom left box
%%
\begin{textblock}{14}(.65, 21.5)
{\large \bfseries Avoiding degeneracy} \\[-.75cm]

For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space. In particular, at $\boldsymbol \theta = \boldsymbol 0$, the RBM is equivalent to elements of $\boldsymbol X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ r.v.s. $\Rightarrow$ No \emph{near-degeneracy}, \emph{instability}, or \emph{uninterpretability}!
\end{textblock}
\begin{textblock}{14.5}(.65, 23.9)
\begin{minipage}{.42\textwidth}
\begin{figure}[ht]
  \begin{minipage}{0.49\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{images/toyhull_top.tikz}

@@

%%-----------------------------------------------
%%
%%Top right box
%%
\begin{textblock}{14}(32.4, 5.71)
{\large \bfseries Manageable (a.k.a. small) examples}
\begin{figure}
\centering
\includegraphics{images/degeneracy.pdf}
\includegraphics{images/instability.pdf}
\includegraphics{images/uninterpretability.pdf}
\label{fig:three_ways}
\caption{As the magnitude of $\boldsymbol \theta$ grows, so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.}
\end{figure}
\end{textblock}

@@

{\bfseries Big takeaway:} RBMs very easily are degenerate, unstable, and uninterpretable. To further complicate things, a rigorous fitting method for these models is not scalable and replicates the nonparametric solution (empirical distribution).
\end{textblock}

>>>>

\begin{table}
\centering
\begin{tabular}{| p{.48\linewidth} | p{.5\linewidth} |}
\hline
Characteristic & Detection \\
\hline
\hline
{\bfseries Near-degeneracy.} Occurs when there is a disproportionate amount of probability placed on only a few elements of the sample space by the model \cite{handcock2003assessing}. & If random variables in the neg-potential function $Q(\cdot)$, having support set $\mathcal{S}$, have a collective mean $\mu(\boldsymbol \theta)$ close to the boundary of the convex hull of $\mathcal{S}$. \\
%If the mean parametrization on the model parameters, $\mu(\boldsymbol \theta)$, is close to the boundary of the convex hull of the set of statistics in the neg-potential function $Q(\boldsymbol x)$. \\
\hline
{\bfseries Instability.} Small changes in natural parameters result in large changes in probability masses, excessive sensitivity \cite{schweinberger2011instability}. & If for any $C > 0$ there exists $N_C > 0$ such that 
$\max\limits_{\boldsymbol x_N \in \mathcal{X}_N}[Q(\boldsymbol x_N)] > CN$ for all $N > N_C$, where $Q(\cdot)$ is the neg-potential function of the model. \\
\hline
{\bfseries Uninterpretability.} Due to the existence of dependence, marginal mean-structure no longer maintained \cite{kaiser2007statistical}. & If the magnitude of the difference between model expectations and expectations under independence (dependence parameters of zero), $\left|\text{E}(\boldsymbol X | \boldsymbol \theta) - \text{E}(\boldsymbol X | \emptyset ) \right|$, is large.\\
\hline
\end{tabular}
\label{tab:degen}
\caption{Table of ``improper model'' characteristics.}
\end{table}
\end{textblock}

%%-----------------------------------------------
%%
%%Bottom left box
%%
\begin{textblock}{14}(.65, 21.5)
{\large \bfseries Data coding to mitigate degeneracy} \\[-.75cm]

For the $\{-1, 1 \}$ encoding of $\mathcal{V}$ and $\mathcal{H}$, the origin is the center of the parameter space. In particular, at $\boldsymbol \theta = \boldsymbol 0$, the RBM is equivalent to elements of $\boldsymbol X$ being distributed as iid Bernoulli$\left(\frac{1}{2}\right)$ $\Rightarrow$ No \emph{near-degeneracy}, \emph{instability}, or \emph{uninterpretability}!
\end{textblock}
\begin{textblock}{14.5}(.65, 23.9)
\begin{minipage}{.42\textwidth}
\begin{figure}[ht]
  \begin{minipage}{0.49\textwidth}
  \resizebox{\linewidth}{!}{
    \tdplotsetmaincoords{60}{-60}
    \input{images/toyhull_top.tikz}

@@

%%-----------------------------------------------
%%
%%Top right box
%%
\begin{textblock}{14}(32.4, 5.71)
{\large \bfseries Manageable (a.k.a. small) examples}
\begin{figure}
\centering
\includegraphics{images/degeneracy.pdf}
\includegraphics{images/instability.pdf}
\includegraphics{images/uninterpretability.pdf}
\label{fig:three_ways}
\caption{As the magnitude of $\boldsymbol \theta$ grows ($\boldsymbol \theta$ moves from $\boldsymbol 0$), so does the occurence of near-degeneracy, instability, and uninterpretability for RBMs of varying sizes.}
\end{figure}
\end{textblock}

@@

{\bfseries Big takeaway:} RBMs very easily are degenerate, unstable, and uninterpretable. As compounding issues, a rigorous fitting method for these models is not scalable and merely replicates the empirical data distribution, a.k.a optimal nonparametric solution.
%only replicates the nonparametric solution (empirical distribution).
\end{textblock}
